---
title: "interactR Vignette"
subtitle: "Conversation Behaviour - Feature Extraction"
author: "Irene Sophia Plank"
date: "`r Sys.Date()`"
format: 
  html:
    fig-width: 8
    fig-height: 5
    toc: true
    number-sections: true
fontsize: 10pt
bibliography: intR.bib
execute:
  warning: false
---

# Introduction

```{r setup}
#| include: false

ls.packages = c(
  "knitr",            # Tables
  "dplyr",
  "ggpubr",           # ggarrange
  "pak"               # to install GitHub packages
)

# if packman is not installed yet, install it
if(!("pacman" %in% installed.packages()[,"Package"])) install.packages("pacman")
pacman::p_load(char = ls.packages) 

# path to the data and the results
dt.path = "2026-08-13_13-50-25"
rs.path = "_preprocessed"
if (!dir.exists(rs.path)) {
  dir.create(rs.path)
}

# set the framerate
fps = 32

# create string of packages
txt = ''
for (package in ls.packages) {
  txt = sprintf("%s%s version %s; ", txt, package, packageVersion(package))
}

redo = F

```

We used ``r R.Version()$version.string`` and the following packages in the creation of this document: ``r txt``

All data was collected in the virtual reality environment VERSE [@verse2026].

The sampling rate of the data is `r fps`. All created datafiles will be saved in ``r rs.path``.

# Setup / installation

```{r}

# for installation, uncomment the following: 
# pak("IreneSophia/interactR")

# load interactR, which contains the preprocessing pipeline, from GitHub
library(interactR)

```

# Extract events

We create a dataframe recording meta-data about this example interaction, including the timepoint of events of interest and VERSE settings. This part needs to be customised to the experiment. In this case, participants were having a conversation based on two slides of questions which were cast to the in-VR screens. After they finished the conversation, we created an eye-tracking visualisation. Screencasts triggered different phases of the event which are logged in the Event file: 

* `Screen: All set to slide S1.PNG` : start of the first round of questions
* `Screen: All set to slide S2.PNG` : start of the second round of questions
* `Screen: All set to slide End.PNG` : end of conversation
* `Screen: All set to slide ET.PNG` : start of the eye tracking visualisation

```{r}

# get all of the events and info from the Event file
df.info = extractEventsVERSE(file.path(dt.path, "EventLog.txt"))

# show the events
kable(df.info)

```

We will save the starting point of the experiment to be able to calculate onset of any Timestamp from the beginning of the experiment. This will allow us to compare the extracted features to a screen recording of the same experiment that we collected. 

```{r}

startExp = df.info |> filter(grepl("started", Event)) |> pull(Timestamp)

```

# Summary of preprocessing pipeline

In this preprocessing pipeline, the following features are extracted from dyadic interactions captured in VERSE: 

* Emotional facial expressions with the emotion categorisation based on @aldenhoven2026real
* Synchrony of joy calculated using windowed-lagged cross-correlations (WLCC), based on @koehler_machine_2024
* Head movements, specifically nodding and head shaking detected with zero crossings
* Synchrony of nodding calculated using wavelet coherence
* Speech and turn-taking features [@plank2023automated], once from VERSE data and once using `praat` [@praat] and the uhm-o-meter [@de_jong_praat_2021]
* Gaze patterns and joint attention

All observed synchrony values are compared to pseudosynchrony values. The assessment of head gestures was inspired by @hale2020you.

Throughout the preprocessing, large data is saved as `arrow` files while aggregated data (one row per interaction partner) is saved as `csv` files. Columns in the `csv` file start with the feature group name, except for dyadic variables which are the same for both interaction partners of the same dyad and have the additional prefix `Dyad`. 

# General data structure

We are going to start by reading in the data from the beginning of the conversation onwards, starting with the presentation of slide `S1.PNG` and ending with the presentation of `ET.PNG` (which we will load separately later).

```{r}

df.info.sel = df.info |> 
  filter(grepl("S1.PNG|ET.PNG", Event)) |>  # focus on the Events of interest
  mutate(
    # let's rename our events to the column names we need to cut out the middle
    Event = if_else(grepl("S1.PNG", Event), "start.use", "end.use")
    ) |>
  # restructure to have one row with the Event as columns
  tidyr::pivot_wider(names_from = Event, values_from = Timestamp)

df = extractDataVERSE(df.info.sel,     # use this events Timestamp as the starting point
                      rs.path = rs.path, # we want the data to be saved in our results folder
                      return = T,        # we also want to dataframe to be returned
                      recompute = redo)  # whether to recompute if it already exists

```

The resulting dataframe is quite large: `r nrow(df)` rows and `r ncol(df)` columns! Let's first have a look at the columns.

```{r}

colnames(df)

```

The columns `Dyad` and `Time` specify the unique VERSE experiment. The column `Identifier` captures the interaction partner. There are columns that capture audio (e.g., `Speaking`), gaze (e.g., `LeftEye_Pos_X`), movement (e.g., `HeadX`) and facial expressions (all columns starting with `Facial_`). The dataframe also contains all the information that was present in the `df.info` in which we captured the events. 

```{r}

kable(head(
  df |> select(Dyad, Time, Identifier, Frame, Avatar) |>
    arrange(Frame)
))

```
When we print the first few rows of the columns describing the interaction partners and the experiment session, we can see that the data is in a mix of wide and long format - each Identifier's data has its own rows but all the different tracked data is captured in columns. 

# Emotional facial expressions

First, we are interested in the facial expressions. We want to extract both aggregated facial expressions associated with emotions and also plot the timecourse of these expressions across the conversation. We can prepare the data with the `featEFE` function. While we could provide our own categorisation of which facial expressions should be captured in which emotional expression, we will use the precoded categorisation based on @aldenhoven2026real.

We are also going to use the `rescaleVERSE` switch: VERSE combines several weighted blendshapes into facial expressions, sometimes resulting in a maximum above 1. Rescaling results in each facial expression being tracked on a scale from 0 to 1. 

```{r}

# use the function
df.efe = featEFE(df,                 # dataframe with tracked VERSE data
                 rescaleVERSE = T,
                 rs.path = rs.path,  # path to the results folder
                 return = T,         # return the aggregated summary
                 recompute = redo)  # whether to recompute if it already exists

kable(df.efe)

```

The aggregated dataframe captures the average emotional facial expressions and the overall facial expressiveness `FEX`, in total but also separately for different `Communication` events - speaking, listening and silence. This wasn't a very emotional conversation, so we don't expect high values here. Let's have a look at the timecourse data that was automatically saved in our results folder. 

```{r}

df.face = arrow::read_feather(file.path(rs.path, "dataEFE.arrow")) |>
  mutate(
    Timecourse = difftime(Timestamp, startExp, units = "min")
  )

```


```{r, echo = F}

# create line plots
df.face |> 
  # drop the Facial columns and focus on the extracted emotional facial expressions
  select(Identifier, Frame, Timecourse, starts_with("EFE"), -EFE_FEX) |>
  tidyr::pivot_longer(cols = starts_with("EFE_"), names_to = 'Emotion',
                      values_to = "EFE") |> 
  # focus on a small part of the data
  filter(Frame >= 1920 & Frame < 3840) |>
  ggplot(aes(x = as.numeric(Timecourse), y = EFE, colour = Emotion)) +
  facet_wrap(. ~ Identifier, nrow = 2) + 
  geom_line(alpha = 0.6, linewidth = 1.2) + theme_bw() + 
  scale_color_brewer(palette = "Dark2") + 
  theme(legend.position = "bottom")

```

We can see that there is bursts of Joy expression, for instance around the 2 minute 40 second mark. This small period is also observable in the screen capture `expressionJoy.mov`. 

# Interpersonal synchrony of joy

In the above example, P1 was smiling before P2 started smiling. This sort of adaptation of facial expressions is something we can capture using interpersonal synchrony (IPS) calculated with windowed-lagged cross-correlation (WLCC). In this technique, we are sliding windows along our time course and shifting them for one interaction partner before correlating the values. Thus, the correlation captures the association between the timecourse data of both interaction partners in these windows, either simultaneously or with one partner "leading" the other, i.e. their behaviour happening first. 

We can calculate the WLCC of Joy using the `extractIPS` function. 

```{r}

df.wlcc = extractIPS(df.face,
                     "EFE_Joy",            # column for which IPS is extracted
                     "WLCC",               # type of IPS
                     fps,                  # frame rate of the data
                     seed = 4480,          # setting a seed for reproducibility
                     rs.path = rs.path,    # results folder
                     winSample = fps * 7,  # size of the window in samples - 7 seconds
                     incSample = fps / 2,  # size of the increment for sliding in samples
                     lagSample = fps * 2,  # size of the lag, i.e., shifting one partner's data in samples
                     recompute = redo)     # whether to recompute if it already exists

kable(head(df.wlcc))

```

The resulting dataframe contains the WLCC values for each lag and window, while also capturing other information for instance what the lag means in the column `Leading`, the seed that was used and which iteration this data comes from. Those last two columns are not relevant when we compute real, observed IPS as also captured in column `Method`, but will become relevant later. 

Let's now visualise the same part of the conversation as above but focusing on Joy. We highlight one window with a high and one window with a low average WLCC value across lags. 

```{r, echo = F}
#| fig.height: 3

df.face |> 
  # drop the Facial columns and focus on the extracted emotional facial expressions
  select(Identifier, Frame, Timecourse, starts_with("EFE"), -EFE_FEX) |>
  # focus on a small part of the data
  filter(Frame >= 1920 & Frame < 3840) |>
  ggplot(aes(x = Frame, y = EFE_Joy, colour = Identifier)) +
  annotate("rect", xmin = 2689, xmax = 3041, ymin = -Inf, ymax = Inf, 
           fill = "lightblue", alpha = 0.4) +
  annotate(
    "text",
    x = (2689 + 3041) / 2, y = 0.85, label = "mean WLCC = 0.57"
  ) +
  annotate("rect", xmin = 2001, xmax = 2353, ymin = -Inf, ymax = Inf, 
           fill = "grey", alpha = 0.4) +
  annotate(
    "text",
    x = (2001 + 2353) / 2, y = 0.85, label = "mean WLCC < 0.01"
  ) + ylim(0, 1) + 
  geom_line(linewidth = 1.2) + theme_bw() + 
  scale_color_brewer(palette = "Dark2") + 
  theme(legend.position = "bottom")

```

We can aggregate the WLCC values using `featWLCC`. 

```{r}

df.wlcc.agg = featWLCC(df.wlcc, 
                       absolute = F,      # we want to keep the signed values
                       r2z = F,           # stick with correlation coefficients
                       rs.path = rs.path,
                       recompute = redo)

kable(df.wlcc.agg)

```

The dataframe contains leading of each interaction partner, prefix = `WLCC`, as well as overall IPS values across all lags with prefix `DyadWLCC`. For each value, the data was aggregated using the grand mean, median, standard deviation as well as the mean of the peak, i.e. maximum value, of each window. While P1 was more likely to lead a joyful expression, the peak is more similar across both interaction partners, suggesting that if P2 was leading a joyful expression it was with a similar association as P1, but they did so less often. 

## Comparison with pseudo IPS

We might be interested in whether the correlations are just artefacts or meaningful adaptation by our interaction partners. To assess this, we can compute pseudo IPS by shuffling either the dyads or the timecourse data of one interaction partner in segments. Since we only have one dyad in our example dataset, we will use segment shuffling and then compare the lag values using permutation testing. 

To create pseudo IPS, we can use the same function as before but not set the number of segment shuffling to 400. Depending on your computer, this may take a few minutes. On a MacBook Air with an M4, shuffling 400 times took about 6min.

```{r}

df.pwlcc = extractIPS(df.face,
                     "EFE_Joy",            # column for which IPS is extracted
                     "WLCC",               # type of IPS
                     fps,                  # frame rate of the data
                     seed = 4480,          # setting a seed for reproducibility
                     rs.path = rs.path,    # results folder
                     nSegment = 400,       # shuffle the segments 400 times per dyad
                     winSample = fps * 7,  # size of the window in samples - 7 seconds
                     incSample = fps / 2,  # size of the increment for sliding in samples
                     lagSample = fps * 2,  # size of the lag, i.e., shifting one partner's data in samples
                     recompute = redo)     # whether to recompute if it already exists

kable(head(df.pwlcc))

```

The resulting dataframe looks exactly the same as the `df.wlcc` from before, but the Method value is `pseudoSegment`. Let's compare whether there is a difference between observed and pseudo IPS of Joy using the function `compareIPS`. While we could do this by ignoring the sign and taking the absolute value of IPS, we want to actually be able to distinguish whether interaction partners react to each other with a similar or the opposite behaviour. 

```{r}

df.comp = compareIPS(rbind(df.wlcc, df.pwlcc),  # bind the two dataframes together as input
                     perm = T,                  # we use permutation testing instead of t-tests
                     alpha = 0.05,
                     rs.path = rs.path,
                     recompute = redo)

kable(head(df.comp))

```

The resulting dataframe contains one row per lag. For each lag, we get the average and standard deviation across the windows, the probability of a difference, the `Direction` (greater or lesser) and `Evaluation` which is `*` if the probability exceeded our alpha. 

We can plot these results with `plotWLCCcomp`. 

```{r}
#| fig.height: 3
#| fig.width: 5

plotWLCCcomp(df.comp, ncol = 1)

```

This plot shows that observed IPS differed significantly from pseudo IPS only for a small block of the positive lags. Specifically, observed IPS exceeds pseudo IPS when P1 was leading, suggesting that P2 reacted to P1's joyful expressions with a more joyful expression as well. This corresponds to what we saw above when plotting the timecourse of joyful expressions as well. 

# Head shaking and nodding

Now, we want to use a Z-crossing algorithm inspired by @hale2020you to detect head shaking and nodding during our conversation. First, we need to preprocess our head movement data: decentring the data and removing artificial phase-wrapping jumps (e.g., flipping between -180 and +180 or 0 and 360 degrees) from rotational values

```{r}

df.head = preproHead(df,
                     c("HeadCenterRotX", "HeadCenterRotY"), # names of the translational columns
                     c("HeadCenterX", "HeadCenterY"),       # names of the rotational columns
                     rs.path = rs.path, 
                     return = T,
                     recompute = redo)

kable(head(df.head))

```

This function saved a new data file, containing only the relevant head movement data and some general information, i.e. `Communication`. We can use this dataframe to detect head gestures (nodding and shaking) using `featHeadGestures`. Here, we are using a minimum framewise degree of movement and filter anything below it. This threshold depends on your setup and your frame rate. To determine it, it can be helpful to create example data, screen capture your movements and compare the results of different thresholds. There are also quite a few additional settings that you can test, as described below in the comments. 

```{r}

df.gesture = featHeadGestures(df.head, 
                              "HeadCenterRotX_fixCirc_detrended", # column for nodding
                              "HeadCenterRotY_fixCirc_detrended", # column for shaking
                              fps, 
                              0.4,   # minimum framewise degree of movement
                              win = 2, # consider windows of length 2 seconds
                              minFreq = 1.5, # focus on movements with at least 1.5Hz
                              maxFreq = 6.5, # and below 6.5Hz
                              winCentre = NULL, # use win for the window size of the centring
                              winSmooth = 0,    # do not use any smoothing
                              rs.path = rs.path, 
                              return = T, 
                              recompute = redo)

kable(head(df.gesture |> select(Identifier, Frame, !any_of(colnames(df.head)))))

```

This table focuses on the newly created column. `Nodding` and `Shaking` contain the preprocessed signal and `diff` columns the framewise change in degree, while columns ending in `zc` track the Zero-crossings. All Zero-crossings above the minimum framewise degree of movement are summed up with a sliding 2-second window, which is captured in the columns ending in `sum`. These sums are only considered relevant head gestures, as captured in the columns ending in `rel`, when they are within the defined frequency range. If head shaking and nodding is detected at the same time, only the gesture that is associated with the larger movement is considered relevant. 

We can now plot the head gestures across time to see where some nodding or head shaking was detected. For this, we can use the plotting function `plotHeadGestures` which plots one Identifier at a time.

```{r}
#| fig-height: 10

minFrame = 1 

p1 = plotHeadGestures(df.head |> filter(Identifier == "P1") |>
                mutate(# adding the timecourse to ease comparison with the screen capture
                  Timecourse = difftime(Timestamp, startExp, units = "secs")
                ), 
                "HeadCenterRotX_fixCirc_detrended", # column for nodding
                "HeadCenterRotY_fixCirc_detrended", # column for shaking
                fps, 
                0.4,   # minimum framewise degree of movement
                minFrame = minFrame,          # where to start plotting
                maxFrame = minFrame + 60*fps, # where to end plotting
                win = 2, # consider windows of length 2 seconds
                minFreq = 1.5, # focus on movements with at least 1.5Hz
                maxFreq = 6.5, # and below 6.5Hz
                winCentre = NULL, # use win for the window size of the centring
                winSmooth = 0,    # do not use any smoothing
                legend = F        # no legend
                )


p2 = plotHeadGestures(df.head |> filter(Identifier == "P2") |>
                mutate( # adding the timecourse to ease comparison with the screen capture
                  Timecourse = difftime(Timestamp, startExp, units = "secs")
                ),  
                "HeadCenterRotX_fixCirc_detrended", # column for nodding
                "HeadCenterRotY_fixCirc_detrended", # column for shaking
                fps, 
                0.4,   # minimum framewise degree of movement
                minFrame = minFrame,          # where to start plotting
                maxFrame = minFrame + 60*fps, # where to end plotting
                win = 2, # consider windows of length 2 seconds
                minFreq = 1.5, # focus on movements with at least 1.5Hz
                maxFreq = 6.5, # and below 6.5Hz
                winCentre = NULL, # use win for the window size of the centring
                winSmooth = 0     # do not use any smoothing
                )

ggarrange(p1, p2, nrow = 2, heights = c(0.44, 0.56))

```

Z-crossings with framewise head movement above the threshold are shown with vertical dashed lines. The solid lines show the preprocessed signal. The columns are the sums of the Z-crossings above the framewise movement threshold, with darker colours for sums within the frequency range. For each interaction partner, nodding and head shaking is plotted separately. We can see that at the almost 2min mark, both interaction partners are using head gestures, with P1 nodding and P2 shaking their head. Again, we can see this nicely in the screen capture `headGestures.mov`. 

# Interpersonal synchrony of nodding

We might also be interested in how aligned people are in their head gestures, specifically nodding. Since these are rhythmic movements, instead of using correlations, we can use wavelet coherence (WTC) which considers both the time and the frequency space. For this, we will use the `biwavelet` package inside our `extractIPS` function. 

```{r}

df.wtc = extractIPS(df.head,
                    "HeadCenterRotX_fixCirc_detrended", # we use this preprocessed column for the IPS
                    "WTC", # instead of WLCC, we want to extract WTC now
                    fps, 
                    featname = "HeadNodding", # since the column name is so long, we rename it
                    rs.path = rs.path, 
                    seed = 8242, 
                    recompute = redo)

kable(head(df.wtc))

```

The resulting dataframe is similar to the WLCC IPS dataframe, but differs in many columns as well. The columns `Frequency` and `Period` track the frequency domain, while `Timecourse` captures the time domain. `COI` gives us the threshold for the cone of influence and `WithinCOI` tracks whether this datapoint is within this cone of influence. `Rsq` captures the coherence, which is similar to a correlation coefficient and captures the strength of the similarity. `Phase` captures whether the relationship is in-phase, anti-phase or driven by one interaction partner leading the other. 

Let's start with plotting the phase distribution with the function `plotPhaseRose`. 

```{r}

plotPhaseRose(df.wtc)

```

The plot shows that P1 was leading more often and that in-phase coherence was more common than anti-phase coherence. We can also load the `biwavelet` object of the dyad that was automatically saved in the results folder and use the plotting function from the `biwavelet` package to investigate this further. 

```{r}

wtc = readRDS(file.path(rs.path, "dataWTC_HeadNodding_P1-P2_seed-8242.rds"))

biwavelet::plot.biwavelet(wtc, plot.phase = T)

```

For some information on how to read this plots, you can check out this tutorial: https://rpubs.com/ibn_abdullah/rwcoher

For IPS based on WTC, we could again use the same functions to create and compare pseudo IPS. However, shuffling segments does not make sense when we are assessing rhythmic behaviour. Dyad shuffling would be preferable, since it preserves the rhythmic nature of the head movements. Since we only have one dyad, we will skip the comparison with pseudo IPS for this vignette. 

# Speech and turn-taking features

## VERSE tracked data

If we don't have any audio recordings, we can extract some speech and turn-taking features based on the sound that was detected by VERSE and included in the tracked data log. We use the function `featSpeech` without setting the `praat.path`. 

```{r}

df.speechV = featSpeech(df, 
                        praat.path = c(), 
                        praat.prefix = c(), 
                        rs.path = rs.path, 
                        return = T, 
                        recompute = redo) 

kable(head(df.speechV))

```

Again, dyadic features in the resulting feature dataframe start with `Dyad`, while individual ones start with `SPCH`. Based on the VERSE data alone, we can extract the number of turns (`DyadSPCH_noTurns`), the median and standard deviation of turn-taking gaps (`SPCH_TurnGapsMedian`, `SPCH_TurnGapsSD`). We also get the time each interaction partner was speaking with `SPCH_PhonationDuration`, as well as the silence-to-turn ratio capturing how much speech versus silence the conversation had (`DyadSPCH_SilenceToTurn`) and the `SPCH_PhonationBalance` which captures who was speaking more. If both interaction partners were speaking a similar amount of the time, then the phonation balance would be 1 for both of them. 

The function also saved a file to the results folder that contains all turns that were detected. We can also plot their distribution. 

```{r}

df.turns = arrow::read_feather(file.path(rs.path, "dataTurn.arrow")) |>
  filter(!is.na(TTG) & abs(TTG) < 6)
  
df.turns |> tidyr::drop_na() |>
  ggplot(aes(x = TTG, fill = Identifier)) + 
  geom_density(alpha = 0.6, colour = NA) + 
  geom_vline(xintercept = 0) + 
  geom_vline(xintercept = median(df.turns$TTG), linetype = "dashed") + 
  geom_label(data = data.frame(xpos = median(df.turns$TTG), ypos = 0.1), size = 2.5, 
             aes(x = xpos, y = ypos, 
                 label = sprintf("Median: %.3fs", median(df.turns$TTG))), fill = "white") + 
  theme_bw() + 
  scale_fill_brewer(palette = "Dark2") + 
  theme(legend.position = "bottom", 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank())

```

We see that gaps between turns were sometimes negative, indicating that P1 and P2's speech was overlapping, a similar finding to previous studies. Before plotting, we filtered any extrem (more than 6 seconds) gaps in either direction. 

## `praat` output

We can now also do the same, but based on the information extracted with the `praat` script. Here, we will only process a small part of the conversation which we will share similarly to the screen capture. Running the `praat` script created four files:

* `Recording_Participant_P1.TextGrid` : containing syllables and speaking versus silent
* `Recording_Participant_P2.TextGrid` : same for other interaction partner
* `intR_pitchIntensity.csv` : pitch and intensity values for all interaction partners
* `intR_syllableNuclei.csv` : aggregated syllable values for all interaction partners

We start our processing by converting the `TextGrid` files into one file readable for R using the function `convertGrid`. 

```{r}

# get a list of all the TextGrid files
ls.files = list.files(path = dt.path, 
                      pattern = ".*.TextGrid",
                      full.names = T)

df.uhm = convertGrid(ls.files, 
                     rs.path = rs.path, 
                     extract = F, # we will extract Dyad and Identifier later
                     return = T,
                     recompute = redo)
```

This dataframe does not contain the Dyad and the Identifier information because `praat` only has the sound files as input. However, we can extract the Identifier from the file name and then add the Dyad and Time information from our main VERSE dataframe. We need this for further processing.

```{r}

# extract Identifier here and then merge with information
df.uhm = df.uhm |>
  mutate(
    Identifier = sub(".*Recording_Participant_(.+).*.TextGrid", "\\1", Path)
  ) |> select(-Path) |>
  # adding in the additional information
  left_join(df |> select(Dyad, Time, Identifier) |> unique(), by = "Identifier")

kable(head(df.uhm))

```

The resulting dataframe contains a list of all the sounding instances of both interaction partners, with their `Start`, `End`, `Duration` and the number of syllables spoken in `nSyll`. If we wanted to, we could now update or create `Speaking`, `Listening` and `Communication` columns in the main dataframe with the tracked VERSE data using the function `addCommunication`; however, we don't need this for this vignette. Thus, our next step is to again aggregate speech and turn-taking features, but now based on the information from `praat`.  

```{r}

# extract speech and turn-taking features
df.speechP = featSpeech(df.uhm, 
                        praat.path = dt.path,  # where can we find the praat files?
                        praat.prefix = "intR", # what prefix was defined in the praat script?
                        rs.path = rs.path, 
                        suffix = "_praat",     # using a suffix to distinguish from the above created feature file
                        return = T, 
                        recompute = redo) 

```
Let's have a look at the two dataframes, but keep in mind that the one produced based on VERSE includes the full conversation while the one based on `praat` only considers part of the conversation (otherwise the wav files would have been to large for GitHub).

```{r}

kable(bind_rows(df.speechV |> select(-Dyad) |> mutate(Source = "VERSE") |> relocate(Source), 
                df.speechP |> select(-Dyad, -Time) |> mutate(Source = "praat") |> relocate(Source)))

```

This dataframe has the same information as the one extracted from the VERSE tracked data, but also the following additional columns: `SPCH_PitchSD` and `SPCH_IntensitySD` capturing the variance of pitch and intensity across the conversation and `SPCH_ArticulationRate` capturing how many syllables per second each Identifier produced. 

# Gaze patterns and joint attention

We start by reading in the data that started with the event of presenting the ET.PNG slide. From this point until the end of the experiment, the participants' eye movements were visualised and we can assess how accurate the gaze tracking was. 

```{r}

df = extractDataVERSE(df.info |> 
                        filter(grepl("ET.PNG", Event)) |>   # focus on the Event of interest
                        mutate(start.use = Timestamp),     # use this events Timestamp as the starting point
                      rs.path = rs.path, # we want the data to be saved in our results folder
                      return = T,        # we also want to dataframe to be returned
                      suffix = "_ET",    # add the suffix ET to the data file
                      recompute = redo)  # whether to recompute if it already exists

```

The dataframe `df` now contains all the data from the Event onwards until the end of the VERSE experiment. We now want to focus on the columns relevant for the gaze patterns. Let's have a look at the first few rows. 

```{r, echo = F}

kable(head(df |> 
  select(Frame, starts_with("AOI") | contains("Target")), 
  n = 2))

```

VERSE automatically tracks when interaction partners' eyes (`LeftEye_Targets` and `RightEye_Targets`) or their head orientation (`Gaze_Targets` in VERSE) hit prespecified targets, for instance the Desk. These columns capture all targets that are hit, even those that are behind another target. The targets are ordered in closeness to the eyes or head. The targets are reported with their respective hit coordinates. Reading in the data with `extractDataVERSE` already extracted the closest target and replaced the interaction partner's Identifier with "Self" versus "Other". These extracted values are captured in `AOI.gaze` for the head orientation and `AOI.left` as well as `AOI.right` for the eye fixations. AOI is the abbreviation for Area of Interest. 

We can use the function `featDwell` to create a data file with periods of fixations as well as an aggregated feature file (CSV) capturing how long each interaction partner spent looking at a specific AOI. The function takes a list of AOIs as input, and all other targets are disregarded. 

```{r}

# list of all the AOIs we want to extract
ls.AOI = c("Other Laptop", "Self Laptop", "Other", "Self", "SharedScreen")

# use the function
df.dwell = featDwell(df,                 # dataframe with tracked VERSE data
                     ls.AOI,             # prespecified list of AOIs
                     rs.path = rs.path,  # path to the results folder
                     return = T,         # return the aggregated summary
                     recompute = redo)  # whether to recompute if it already exists

kable(df.dwell)

```

Individual dwell times start with the prefix `Dwell` and if the dataframe contained information on `Communication`, then they are aggregated both in Total dwell times in percent but also separated for speaking, listening and silence.

The features starting with `DyadDwell` capture joint attention, when both interaction partner are fixating at the same, i.e. the shared screen, or corresponding object, e.g. when they are both looking at each other's heads. Thus, these values are the same for both interaction partners. 

In addition to this feature file, the function also saves the periods of fixations. 

```{r}

df.period = arrow::read_feather(file.path(rs.path, "dataDwell.arrow"))

kable(head(df.period))

```

We want to compare this to our screen recording with visualised eye tracking to assess how well it worked. To do this, we visualise the periods of the screen capture using the function `plotETperiods`. 

```{r, echo = F}

# add the timecourse for easier comparison with screen capture
df.period = df.period |>
  mutate(
    minTimecourse = as.numeric(difftime(minTime, startExp, units = "secs")),
    maxTimecourse = as.numeric(difftime(maxTime, startExp, units = "secs")),
  )

# plot it using the function
plotETperiods(df.period, fps)

```

This shows us that P1 starts with looking at P2, who also looks back around the 409 ~ 6min 49s mark. Then after almost 20 seconds, they both look at the laptops, before moving on to also look at the shared screen. This pattern corresponds to the screen capture `eyeTracking.mov`. 

# References